Altering Backward Pass Gradients to improve Convergence

### 2022-11-17 0100.tex
Mod4
Mod5
Mod6

\begin{abstract}
In standard neural network training, the gradients in the backward pass are determined
by the forward pass. As a result, the two stages are coupled. This is how most neural
networks are trained currently. However, gradient modification in the backward pass has
seldom been studied in the literature. In this paper we explore decoupled training,
where we alter the gradients in the backward pass. We propose a simple yet powerful
method called PowerGrad Transform (PGT), that alters the gradients before the weight
update in the backward pass and significantly enhances the predictive performance of the
neural network. PGT trains the network to arrive at a better optima at convergence. It
is computationally extremely efficient, virtually adding no additional cost to either
memory or compute, but results in improved final accuracies on both the training and
test sets. PGT is easy to integrate into existing training routines, requiring just a
few lines of code. PGT accelerates training and makes it possible for the network to
better fit the training data. With decoupled training, PGT improves baseline accuracies
for ResNet-50 by 0.73\%, for SE-ResNet-50 by 0.66\% and by more than 1.0\% for the
non-normalized ResNet-18 network on the ImageNet classification task.
\keywords{backpropagation \and gradients \and neural networks \and softmax \and clipping.}
\end{abstract}

@@

where first the gradient $\nabla_{W_D}(L)$ is transformed using the transformation
function `$f$' and then this transformed gradient is propagated through the rest of the
backward graph. All other gradient vectors $\nabla_{W_i}(L)^{*}$ are computed as is, but
because of the early injection of the transformed gradient $\nabla_{W_D}(L)$, all other
gradient vectors that are connected to the transformed gradient through the chain rule
($\nabla_{W_i}(L)^{*}$,\ \ i=D-1,\ldots,1), gets subsequently altered.

@@

\item We perform theoretical analysis of the properties of the PowerGrad transformation
(section \ref{sec:pgt_prop}) and explore its effect on weight parameters and gradients,
logits, loss values and class separation (section \ref{sec:pgt_effects}).

@@

Gradient Clipping \cite{pascanu2013difficulty} is a technique that involves changing or
clipping the gradients with respect to a predefined threshold value during backward
propagation through the network and updating the weights using the clipped gradients
\cite{zhang2019gradient}\cite{smith2020generalization}. By rescaling the gradients, the
weight updates are likewise rescaled, significantly reducing the risk of an overflow or
underflow \cite{pascanu2012understanding}. GC can be used for training networks without
batch-normalization. At larger batch sizes, the clipping threshold in GC becomes highly
sensitive and requires extensive finetuning for various models, batch sizes, and
learning rates. As we demonstrate later in our studies, GC is not as effective in
improving the performance of non-normalized networks. Adaptive Gradient Clipping
\cite{brock2021high} is developed to further enhance backward pass gradients than what
is performed by GC. It takes into account the fact that the ratio of the gradient norm
to the weight norm can provide an indication of the expected change in a single step of
optimization. Label smoothing, introduced by Szegedy et al.
\cite{szegedy2016rethinking}, utilizes smoothing of the ground truth labels as a method
to impose regularization on the logits and the weights of the neural network. AGC
performs better than GC in non-normalized networks. However, we show that PGT
outperforms both in such networks.

Knowledge distillation \cite{hinton2015distilling} is a process in which two networks
are trained with hard and soft labels alternatively. Variants of knowledge distillation
include self-distillation \cite{zhang2018deep}, identical student network distillation
\cite{furlanello2018born}, channel distillation \cite{ge2019distilling}, regularizing
wrong predictions \cite{yun2020regularizing} and representation or embedding based
knowledge distillation \cite{yao2018deep}\cite{aguilar2020knowledge}. The key difference between PGT and knowledge
distillation is that in the latter, the transformation is applied in the forward pass,
while PGT is a backward pass modification only. Also, in distillation settings the
temperature parameter is a part of the network's computation graph. In the case of
PowerGrad Transform, we directly tamper the gradients without introducing any change in
the forward pass. PGT differs from self-knowledge distillation as it neither introduces
any additional sub-modules nor creates different ensembles to improve the performance of
the model. PGT follows the standard neural network training mechanism with modified
gradients.

@@

We provide the proofs in the supplementary section and conclude that the stationary
threshold is a monotonic non-decreasing function with respect to $\alpha$. Also the
derivative of PGT function with respect to the true probabilities is non-negative which
in turn means that the transformation is an order-preserving map. All values greater
than the threshold move towards the threshold after transformation and all values below
the threshold also move towards the threshold, and the threshold itself moves
monotonically towards $1/C$ as $\alpha$ is decreased from $1$ to $0$. This concludes
that the transformation smooths the original distribution.

\subsection{Effects of PGT on logit, weight gradients and class separation}
\label{sec:pgt_effects}

PGT, other than having the properties mentioned above, also:
\begin{enumerate}
\item restricts the magnitude of the directional derivative for each logit from becoming too small or too high
\item enhances gradient behavior of weights in a manner that improves learning
\item leads to better class separation.
\end{enumerate}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. With consistent improvements in training accuracies across all cases, we
conclude PGT helps networks learn better representations and arrive at better optimas
during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a, b).
Practioners can also choose to accelerate training and save as much as $40\%$ of the
epoch budget [Fig. \ref{fig:metrics}(c)].

@@

We examine issues that occur in non-normalized networks (networks without BN layers). We
use ResNet-18 \cite{he2016deep} as the foundation model trained on ImageNet-1K
\cite{russakovsky2015imagenet}. Deeper networks such as ResNet-34 and ResNet-50 are
impossible to train without Batch Normalization courtesy of the increased depth and so
we solely focus on ResNet-18. We designate different layers with their corresponding
layer indices. Throughout the training process, we monitor variations in the the
per-filter L2-norm of each layer's weights (layer indices are mentioned in section
\ref{sec:r18_layers}). In Fig. \ref{fig:norm_plots}(a), some filters of layer $11$
achieve a norm of zero during training. We refer to this event as `Zeroing Out', and it
occurs when one of the channels (or filters) of a weight tensor gets fully filled with
zeros and such filters do not contribute at all to determine the input-output
relationship of a dataset, as the feature tensor it produces is also an all-zero tensor.
According to our findings, when a filter once zeroed out does not recover with further
training.

@@

\captionsetup{font=footnotesize}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Norm vs iter. plots demonstrating layer characteristics (the zero out
phenomena) and the efficacy of PGT and AGC+PGT over baseline. Each colour represents a
different filter or feature vector of a particular layer of a non-BN variant of
ResNet-18. }
\label{fig:norm_plots}
\vspace{-0.5cm}
\end{figure}

Fig. \ref{fig:norm_plots}(c) and \ref{fig:norm_plots}(e) are the plot of the final conv
layer's filters (layer $19$) and the features output of global average pooling (GAP)
layer respectively. We observe a number of filters in the final layer completely zeroing
out. The feature vector after GAP layer directly interfaces with the fully connected
layer immediately after. Therefore any zeroed out features leads to permanent
information loss, as it does not contribute to the learning of decision boundaries in
the fully connected layer. Also, zeroed out weights tensors lead to zeroed out gradients
hence stopping training for all subsequent iterations leading to a collapse in training
for the affected layers. With large batch sizes, it is possible that an entire layer
zeroes out [Fig. \ref{fig:high_bs}]. Gradient modification methods such as PGT or AGC
can alleviate the zeroing out phenomena. We observe that the number of zeroed out
filters has considerably reduced [Fig. \ref{fig:norm_plots}(b, d)]. The final feature
tensor [Fig. \ref{fig:norm_plots}(f)] with PGT enabled, does not contain any zeroed out
regions indicating that information loss is mitigated as the features pass on from the
feature extracting layers to the fully connected layer. When PGT and AGC are combined,
we get the most benefit for layer $15$ [Fig \ref{fig:norm_plots}(g-i)] and layer $2$
[Fig \ref{fig:norm_plots}(j-l)].

In Table \ref{tab:wobn_table} we find the baseline performance for high batch sizes
($1024$) is drastically inferior to baselines for other batch sizes. PGT helps regain
some of the lost performance by $0.682\%$ ($65.498\%$ vs. $64.816\%$). At a batch size
of 512, invoking PGT improves the training accuracy baseline by $1.48\%$ and the test
accuracy baseline by $0.684\%$, while at a batch size of 256, the improvement in
training and test accuracies are $1.11\%$ and $1.018\%$ respectively. In comparison, the
test accuracy improvements obtained by GC and AGC at batch size of 256
is much less at $0.27\%$ and $0.5\%$, respectively. On the training accuracy front,
since we get a significant boost ($1.48\%$ at batch size of 512 and $1.11\%$ at batch
size of 256), it leads us to infer that when PowerGrad Transform is used, the network

\begin{wrapfigure}{l}{0.4\textwidth}
\vspace{-1cm}
\centering
\includegraphics[width=0.45\textwidth]{baseline_high_bs-w-layer2-2}
\caption{}
\label{fig:high_bs}
\vspace{-2cm}
\end{wrapfigure}

\noindent
fits the training dataset more tightly and the convergence optima is significantly
superior. Combining GC with PGT is not very significant. When AGC and PGT are combined,
we see a tremendous increase in test accuracy of over $2.06\%$ over the baseline.
Additionally, we provide comprehensive graphs of the norm profiles of the AGC + PGT
method for all layers in the supplementary section \ref{sec:plots1}\ -\
\ref{sec:plots4}.

@@

@@

\begin{figure}[!bp]
\centering
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_alpha}
\caption{\\Logit vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{loss_vs_alpha}
\caption{\\Cross-entropy loss vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_loss}
\caption{\\Logit-Loss correlation}
\end{subfigure}
\caption{ Plots of various statistical measures: \textbf{(a)} Variations in the logit
norm vs. $\alpha$. The logit norm is calculated per image over the test set of
ImageNet-1K (at the linear layer of the ResNet-18 architecture) and then averaged.
\textbf{(b)} Variations in the final loss values obtained for different $\alpha$
settings. \textbf{(c)} Correlation between loss and logits. Regression line equation:
$(0.01074 x + 0.25816)$. }
\label{fig:stats}
\end{figure}

@@

\subsection{Proofs}

@@

\subsection{Norm plots of weights (baseline (low batch size)) of non-BN ResNet-18}
\label{sec:plots1}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{baseline-w_norm}
\caption{ Method: baseline, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

@@

\subsection{Norm plots of weights (baseline (high batch size)) of non-BN ResNet-18}
\label{sec:plots2}
\begin{figure}[ht] \centering
\includegraphics[width=0.57\textwidth]{baseline_high_bs-w_norm} \caption{ Method:
baseline (high batch size), batch size = 1024. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

@@

\subsection{Norm plots of weights (PGT) of non-BN ResNet-18}
\label{sec:plots3}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{pgt-w_norm}
\caption{ Method: PGT, batch size = 256. This is a plot depicting the iteration-wise
evolution of the norm of each filter in each layer. Each subplot corresponds to a layer
of the unnormalized ResNet-18. Above each subplot, the layer indices are shown. Each
filter in each subplot is shown with a separate and randomly determined colour. }
\end{figure}

@@

\subsection{Norm plots of weights (AGC + PGT) of non-BN ResNet-18}
\label{sec:plots4}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{agc_pgt-w_norm}
\caption{ Method: AGC + PGT, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. } \end{figure}

@@

For the layer indices used in figures Fig. \ref{fig:norm_plots}, \ref{fig:high_bs},
\ref{fig:norm_plots}, we provide a block diagram depicting the different convolutional
layers of ResNet-18 with corresponding layer indices in Fig. \ref{fig:resnet18}. Each
layer represents a convolutional layer, with the layer index and the number of filters
denoted alongside it. Downsampling convolutional layers in skip connections are drawn
and their layer indices are shown on the skip connection paths.

@@

>>>>

\begin{abstract}
In standard neural network training, the gradients in the backward pass are determined
by the forward pass. As a result, the two stages are coupled. This is how most neural
networks are trained hitherto. Gradient modification in the backward pass has seldom
been studied in the literature. In this paper we explore decoupled training, where we
alter the gradients in the backward pass. We propose a simple yet powerful method called
PowerGrad Transform (PGT), that alters the gradients before the weight update in the
backward pass and significantly enhances the predictive performance of a convolutional
neural network. PGT trains networks to arrive at a better optima at convergence. It is
computationally efficient, and adds no additional cost to either memory or compute, but
results in improved final accuracies on both the training and test datasets. PowerGrad
Transform is easy to integrate into existing training routines, requiring just a few
lines of code. PGT accelerates training and makes it possible for the network to better
fit the training data. With decoupled training, our method improves baseline accuracies
for ResNet-50 by 0.73\%, for SE-ResNet-50 by 0.66\% and by more than 1.0\% for the
non-normalized ResNet-18 network on the ImageNet classification task.
\keywords{backpropagation \and gradients \and softmax \and clipping \and classification.}
\end{abstract}

@@

where the gradient $\nabla_{W_D}(L)$ is first transformed using the transformation
function `$f$' and then this transformed gradient is propagated through the rest of the
backward graph. All other gradient vectors $\nabla_{W_i}(L)^{*}$ are computed as is, but
because of the early injection of the transformed gradient $\nabla_{W_D}(L)$, all other
gradient vectors that are connected to the transformed gradient through the chain rule
($\nabla_{W_i}(L)^{*}$,\ \ i=D-1,\ldots,1), gets subsequently altered.

@@

\item We perform theoretical analysis of the properties of the PowerGrad transformation
and explore its effect on weight parameters and gradients, logits, loss values and class
separation (section \ref{sec:pgt_prop}).

@@

Gradient Clipping \cite{pascanu2013difficulty} is a gradient modification method that
involves clipping/altering the gradients with respect to a predefined threshold value
during backward propagation through the network and updating the weights using the
clipped gradients \cite{zhang2019gradient}\cite{smith2020generalization}. By rescaling
the gradients, the weight updates are likewise rescaled, significantly reducing the risk
of an overflow or underflow \cite{pascanu2012understanding}. GC can be used for training
networks without batch-normalization. At larger batch sizes, the clipping threshold in
GC becomes highly sensitive and requires extensive finetuning for various models, batch
sizes, and learning rates. As we demonstrate later in our studies, GC is not as
effective in improving the performance of non-normalized networks. Adaptive Gradient
Clipping \cite{brock2021high} is developed to further enhance backward pass gradients
than what is performed by GC. It takes into account the fact that the ratio of the
gradient norm to the weight norm can provide an indication of the expected change in a
single step of optimization. Label smoothing, introduced by Szegedy et al.
\cite{szegedy2016rethinking}, utilizes smoothing of the ground truth labels as a method
to impose regularization on the logits and the weights of the neural network. AGC
performs better than GC in non-normalized networks. However, we show that PGT
outperforms both in networks such as ResNets.

Knowledge distillation (KD) \cite{hinton2015distilling} is a process in which two
networks are trained with hard and soft labels alternatively. Variants of knowledge
distillation include self-distillation \cite{zhang2018deep}, identical student network
distillation \cite{furlanello2018born}, channel distillation \cite{ge2019distilling},
regularizing wrong predictions \cite{yun2020regularizing} and representation or
embedding based knowledge distillation \cite{yao2018deep}. Even though both PGT and KD
requires probability manipulation, the key difference is that in the latter the
transformation is applied in the forward pass, while PGT is a backward pass modification
only. In distillation settings the temperature parameter is a part of the network's
computation graph. In the case of PowerGrad Transform, we directly tamper the gradients
without introducing any change in the forward pass. PGT differs from self-knowledge
distillation as it neither introduces any additional sub-modules nor creates different
ensembles to improve the performance of the model. PGT follows the standard neural
network training mechanism with modified gradients.

@@

We provide the proofs in supplementary section \ref{sec:proofs} and conclude that the
stationary threshold is a monotonic non-decreasing function with respect to $\alpha$.
Also the derivative of PGT function with respect to the true probabilities is
non-negative which in turn means that the transformation is an order-preserving map. All
values greater than the threshold move towards the threshold after transformation and
all values below the threshold also move towards the threshold, and the threshold itself
moves monotonically towards $1/C$ as $\alpha$ is decreased from $1$ to $0$. This
concludes that the transformation smooths the original distribution.

PGT, other than having the properties mentioned above, also satisfies the following. The
transformation:
\begin{enumerate}
\item restricts the magnitude of the directional derivative for each logit from becoming too small or too high
\item enhances gradient behavior of weights in a manner that improves learning
\item leads to better class separation.
\end{enumerate}

@@

In our experiments with Squeeze-and-Excitation variant of ResNet-50 i.e.
SE-ResNet-50\cite{hu2018squeeze}, we observe significant improvements; a $0.97\%$ boost
in training accuracy and a $0.734\%$ increase in test accuracy. In our experiments with
ResNet-50, we find an increase of $0.5\%$ and $0.656\%$ performance enhancement over the
cosine scheduler baseline for training and testing respectively. The corresponding
improvement over the step scheduler baseline for ResNet-50 is $0.57\%$ (training) and
$0.524\%$ (testing). ResNet-101 sees a higher improvement in training fit, $0.81\%$ to
be exact, while the improvement over the test set is $0.362\%$. Smaller networks such as
SE-ResNet-18 and ResNet-18 sees accuracy boosts which are smaller but nevertheless
positive. With consistent improvements in training accuracies across all cases, we
conclude PGT helps networks learn better representations and arrive at better optimas
during convergence. Per epoch training and test accuracy plots of ResNet-18 and
ResNet-50 (both with and without PGT) are shown in Fig. \ref{fig:metrics}(a,b).
Practioners can also choose to accelerate training and save as much as $40\%$ of the
epoch budget [Fig. \ref{fig:metrics}(c)].

@@

We examine issues that occur in non-normalized networks (networks without BN layers). We
use ResNet-18 \cite{he2016deep} as the foundation model trained on ImageNet-1K
\cite{russakovsky2015imagenet}. Deeper networks such as ResNet-34 and ResNet-50 are
impossible to train without Batch Normalization courtesy of the increased depth and so
we solely focus on ResNet-18. We designate different layers with their corresponding
layer indices. Throughout the training process, we monitor variations in the the
per-filter L2-norm of each layer's weights (layer indices are mentioned in section
\ref{sec:r18_layers}). In Fig. \ref{fig:norm_plots}(a), some filters of layer $11$
achieve a norm of zero during training. We refer to this event as `Zeroing Out', and it
occurs when one of the channels (or filters) of a weight tensor gets fully filled with
zeros and such filters do not contribute at all to determine the input-output
relationship of a dataset, as the feature tensor it produces is also filled with zeros
for the corresponding filter. When a filter once zeroes out, it does not recover with
further training, as all gradients that it receives in future iterations are all zeros.

@@

\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.16\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\captionsetup{font=normalsize}
\caption{ Norm vs iter. plots demonstrating layer characteristics (the zero out
phenomena) and the efficacy of PGT and AGC+PGT over baseline. Each colour represents a
different filter or feature vector of a particular layer of a non-BN variant of
ResNet-18. Readers are requested to zoom in or refer to section \ref{sec:big_plots} for
a zoomed-in view. }
\label{fig:norm_plots}
\vspace{-0.5cm}
\end{figure}

Fig. \ref{fig:norm_plots}(c) and \ref{fig:norm_plots}(e) are plots of the final conv
layer's filters (layer $19$) and the features output of final global average pooling
(GAP) layer respectively. We observe a number of filters and features in the final layer
completely zeroing out. The feature vector after the GAP layer directly interfaces with
the fully connected layer. Therefore any zeroed out features leads to permanent
information loss, as it does not contribute to the learning of decision boundaries in
the fully connected layer. Also, zeroed out weights tensors lead to zeroed out gradients
hence stopping training for all subsequent iterations leading to a collapse in training
for the affected layers. With large batch sizes, it is possible that an entire layer
zeroes out as shown in the figure below. Gradient modification methods such as PGT or
AGC can alleviate the zeroing out phenomena as we observe that the number of zeroed out
filters has considerably reduced [Fig. \ref{fig:norm_plots}(b, d)]. The final feature
tensor [Fig. \ref{fig:norm_plots}(f)] with PGT enabled, does not contain any zeroed out
regions indicating that information loss is mitigated as the features pass on from the
feature extracting layers to the fully connected layer. When PGT and AGC are combined,
we get the most benefit as we demonstrate for layers $15$ and layer $2$ (Fig
\ref{fig:norm_plots}(g-i)] and Fig \ref{fig:norm_plots}(j-l) respectively).

@@

In Table \ref{tab:wobn_table} we find that the baseline performance for high batch sizes
($1024$) is drastically inferior to baselines for other batch sizes. PGT helps regain
some of the lost performance by $0.682\%$ ($65.498\%$ vs. $64.816\%$). At a batch size
of 512, invoking PGT improves the training accuracy baseline by $1.48\%$ and the test
accuracy baseline by $0.684\%$, while at a batch size of 256, the improvement in
training and test accuracies are $1.11\%$ and $1.018\%$ respectively. In comparison, the
test accuracy improvements obtained by GC and AGC at batch size of 256

\begin{wrapfigure}{l}{0.4\textwidth}
\vspace{-1cm}
\centering
\includegraphics[width=0.45\textwidth]{baseline_high_bs-w-layer2-2}
\vspace{-1.5cm}
\end{wrapfigure}

\noindent
is much less at $0.27\%$ and $0.5\%$, respectively. On the training accuracy front,
since we get a significant boost ($1.48\%$ at batch size of 512 and $1.11\%$ at batch
size of 256), it leads us to infer that when PowerGrad Transform is used, the network
fits the training dataset more tightly and the convergence optima is significantly
superior. Combining GC with PGT is not very significant. When AGC and PGT are combined,
we see a tremendous increase in test accuracy of over $2.06\%$ over the baseline.
Additionally, we provide comprehensive graphs of the norm profiles of the AGC + PGT
method for all layers in the supplementary section \ref{sec:plots1}\ -\
\ref{sec:plots4}.

@@

\begin{figure}[!t]
\centering
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_alpha}
\caption{\\Logit vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{loss_vs_alpha}
\caption{\\Cross-entropy loss vs. $\alpha$}
\end{subfigure}%
\begin{subfigure}{0.33\textwidth}
\centering
\includegraphics[width=0.98\textwidth]{logit_vs_loss}
\caption{\\Logit-Loss correlation}
\end{subfigure}
\caption{ Plots of various statistical measures: \textbf{(a)} Variations in the logit
norm vs. $\alpha$. The logit norm is calculated per image over the test set of
ImageNet-1K (at the linear layer of the ResNet-18 architecture) and then averaged.
\textbf{(b)} Variations in the final loss values obtained for different $\alpha$
settings. \textbf{(c)} Correlation between loss and logits. Regression line equation:
$(0.01074 x + 0.25816)$. }
\label{fig:stats}
\vspace{-0.5cm}
\end{figure}

@@

\subsection{Proofs}
\label{sec:proofs}

@@

\subsection{Norm plots of weights (baseline (low batch size)) of non-BN ResNet-18}
\label{sec:plots1}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{baseline-w_norm}
\caption{ Method: baseline, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. Readers are requested to zoom in. } \end{figure}

@@

\subsection{Norm plots of weights (baseline (high batch size)) of non-BN ResNet-18}
\label{sec:plots2}
\begin{figure}[ht] \centering
\includegraphics[width=0.57\textwidth]{baseline_high_bs-w_norm} \caption{ Method:
baseline (high batch size), batch size = 1024. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour. Readers are requested to zoom in. } \end{figure}

@@

\subsection{Norm plots of weights (PGT) of non-BN ResNet-18}
\label{sec:plots3}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{pgt-w_norm}
\caption{ Method: PGT, batch size = 256. This is a plot depicting the iteration-wise
evolution of the norm of each filter in each layer. Each subplot corresponds to a layer
of the unnormalized ResNet-18. Above each subplot, the layer indices are shown. Each
filter in each subplot is shown with a separate and randomly determined colour. Readers
are requested to zoom in. }
\end{figure}

@@

\subsection{Norm plots of weights (AGC + PGT) of non-BN ResNet-18}
\label{sec:plots4}
\begin{figure}[ht] \centering \includegraphics[width=0.57\textwidth]{agc_pgt-w_norm}
\caption{ Method: AGC + PGT, batch size = 256. This is a plot depicting the
iteration-wise evolution of the norm of each filter in each layer. Each subplot
corresponds to a layer of the unnormalized ResNet-18. Above each subplot, the layer
indices are shown. Each filter in each subplot is shown with a separate and randomly
determined colour.
Readers are requested to zoom in. } \end{figure}

@@

For the layer indices used in figures Fig. \ref{fig:norm_plots}, \ref{fig:norm_plots},
we provide a block diagram depicting the different convolutional layers of ResNet-18
with corresponding layer indices in Fig. \ref{fig:resnet18}. Each layer represents a
convolutional layer, with the layer index and the number of filters denoted alongside
it. Downsampling convolutional layers in skip connections are drawn and their layer
indices are shown on the skip connection paths.

@@

\subsection{Plot of Figure \ref{fig:norm_plots}}
\label{sec:big_plots}

\begin{figure}[h]
\centering

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-4-2}
\caption{Layer 11\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-4-2}
\caption{Layer 11\\ \forceindentb\textbf{PGT}}
\end{subfigure}

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-7-2}
\caption{Layer 19\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-7-2}
\caption{Layer 19\\ \forceindentb\textbf{PGT}}
\end{subfigure}

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-f-layer-22-1}
\caption{GAP features\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-f-layer-22-1}
\caption{GAP features\\ \forceindentb\textbf{PGT}}
\end{subfigure}
\caption{ Norm vs iter. plots demonstrating layer characteristics (the zero out
phenomena). Each colour represents a different filter or feature vector of a particular
layer of a non-BN variant of ResNet-18. }
\end{figure}

\begin{figure}[h]
\centering

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindentb\textbf{PGT}}
\end{subfigure}

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/baseline-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{baseline}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindentb\textbf{PGT}}
\end{subfigure}

\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-5-3}
\caption{Layer 15\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\begin{subfigure}[t]{0.34\textwidth}
\includegraphics[width=\textwidth]{trimmed/agc_pgt-w-layer-1-2}
\caption{Layer 2\\ \forceindenta\textbf{AGC+PGT}}
\end{subfigure}
\caption{ Norm vs iter. plots demonstrating the efficacy of PGT and AGC+PGT over
baseline. Each colour represents a different filter or feature vector of a particular
layer of a non-BN variant of ResNet-18. }
\end{figure}
